---
title: "data_wrangling"
author: "Christopher SHerve"
date: "2026-04-14"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(knitr)
```

## Step 1 — Load Raw Data Files

Four raw Excel files are loaded. Each file required different parsing
due to inconsistent formatting across lab report templates. We first
examine the structure of each file before cleaning.

```{r load-yield}
# Yield data — primary source for plot numbers, treatments, and agronomic measurements
yield_raw <- read_excel("data/raw/Yield.xlsx", skip = 1)
glimpse(yield_raw)
```

## Step 2 — Clean Yield Data

The yield file is the primary source for plot numbers, treatment codes,
block structure, and agronomic measurements. Column names were generic
due to Excel formatting so columns are selected by position and renamed
explicitly. Treatment codes were verified against lysimeter and soils
files with all 32 plots confirmed across both years.

```{r clean-yield}
yield_clean <- yield_raw %>%
  dplyr::select(
    year            = 1,
    location        = 2,
    plot            = 3,
    replicate       = 4,
    trt             = 5,
    crops           = 6,
    legume_inclusion = 7,
    n_fertilizer    = 8,
    grain_yield     = 9,
    iwg_straw_yield = 10,
    legume_biomass  = 11,
    plant_height    = 12,
    lodging_score   = 13
  ) %>%
  filter(!is.na(plot)) %>%
  mutate(
    year            = as.numeric(year),
    plot            = as.numeric(plot),
    trt             = as.numeric(trt),
    n_fertilizer    = as.numeric(n_fertilizer),
    grain_yield     = as.numeric(grain_yield),
    iwg_straw_yield = as.numeric(iwg_straw_yield),
    legume_biomass  = as.numeric(legume_biomass),
    location        = case_when(
      plot < 2000  ~ "Waseca",
      plot >= 2000 ~ "Lamberton")
  )

glimpse(yield_clean)

# Verify treatment codes
yield_clean %>%
  distinct(trt, crops, n_fertilizer) %>%
  arrange(trt) %>%
  kable(caption = "Treatment code verification")
```

## Step 3 — Load and Inspect Lysimeter Data

The lysimeter file contains two sheets with different formats — 2014
data from Agvise Laboratories and 2015 data from Research Analytical
Laboratory (UMN). Each sheet requires separate parsing.

```{r inspect-lysimeter}
# Check 2014 sheet structure
lys_2014_raw <- read_excel("data/raw/Lysimeter.xlsx", 
                            sheet = "2014", skip = 14)
glimpse(lys_2014_raw)
```

```{r}
### 3a — Clean 2014 Lysimeter Data

#### The 2014 lab reported below detection limit (BDL)values as numeric zero. Sampling occurred in May, June, July, September, and October. August was not sampled — coinciding with grain harvest operations (mid-August; Jungers et al. 2017).

lys_2014_clean <- lys_2014_raw %>%
  dplyr::select(
    plot    = `1105`,
    month   = `May`,
    NO3_raw = `0.628`
  ) %>%
  filter(!is.na(plot), !is.na(NO3_raw)) %>%
  mutate(
    plot    = as.numeric(plot),
    NO3_raw = as.numeric(NO3_raw),
    year    = 2014,
    is_bdl  = NO3_raw == 0,
    NO3_sub = if_else(is_bdl, 0.05, NO3_raw)
  ) %>%
  filter(!is.na(NO3_sub))

# Check months present
lys_2014_clean %>%
  distinct(month) %>%
  kable(caption = "2014 lysimeter sampling months")

# BDL summary
lys_2014_clean %>%
  summarise(
    n_total = n(),
    n_bdl   = sum(is_bdl),
    pct_bdl = round(mean(is_bdl) * 100, 1)
  ) %>%
  kable(caption = "2014 BDL summary")

#### 2015
lys_2015_raw <- read_excel("data/raw/Lysimeter.xlsx", 
                            sheet = "2015", skip = 13)
names(lys_2015_raw)

lys_2015_clean <- lys_2015_raw %>%
  dplyr::select(
    location  = `Waseca`,
    plot_3    = `125`,
    trt_code  = `I0`,
    NO3_raw   = `< 0.1`,
    date      = `42491`
  ) %>%
  filter(!is.na(plot_3), !is.na(location)) %>%
  mutate(
    plot_3   = as.numeric(plot_3),
    plot     = case_when(
      location == "Waseca"    ~ 1000 + plot_3,
      location == "Lamberton" ~ 2000 + plot_3),
    year     = 2015,
    # Date is already a datetime — just extract month
    month    = format(as.Date(date), "%b"),
    is_bdl   = case_when(
      str_detect(as.character(NO3_raw), "<") ~ TRUE,
      suppressWarnings(!is.na(as.numeric(NO3_raw))) &
        as.numeric(NO3_raw) < 0.1 ~ TRUE,
      TRUE ~ FALSE),
    NO3_sub  = case_when(
      is_bdl ~ 0.05,
      str_detect(as.character(NO3_raw), "/") ~
        map_dbl(str_split(as.character(NO3_raw), "/"),
                ~mean(as.numeric(str_trim(.x)), na.rm = TRUE)),
      TRUE ~ suppressWarnings(as.numeric(NO3_raw)))
  ) %>%
  filter(!is.na(NO3_sub))

# Check months
distinct(lys_2015_clean, month)

lys_2015_clean <- lys_2015_clean %>%
  filter(month != "Apr")

# Verify
distinct(lys_2015_clean, month)
nrow(lys_2015_clean)
lys_2015_clean %>%
  summarise(
    n_total = n(),
    n_bdl   = sum(is_bdl),
    pct_bdl = round(mean(is_bdl) * 100, 1)
  )
```
```{r}
# Combining of lysimeter data ----
# Combine both years
lys_all <- bind_rows(
  lys_2014_clean %>% 
    dplyr::select(plot, year, NO3_sub, is_bdl),
  lys_2015_clean %>% 
    dplyr::select(plot, year, NO3_sub, is_bdl)
)

# Aggregate to plot-year level
lys_summary <- lys_all %>%
  group_by(plot, year) %>%
  summarise(
    NO3_mean_mgL     = mean(NO3_sub, na.rm = TRUE),
    n_raw_readings   = n(),
    n_bdl_readings   = sum(is_bdl),
    pct_bdl_readings = round(mean(is_bdl) * 100, 1),
    any_bdl          = as.numeric(any(is_bdl)),
    majority_bdl     = as.numeric(mean(is_bdl) > 0.5),
    .groups = 'drop'
  )

# Check
nrow(lys_summary)
head(lys_summary, 10)

```

```{r}
# Moisture----
moisture_raw <- read_excel("data/raw/Moisture.xlsx")
names(moisture_raw)
nrow(moisture_raw)
head(moisture_raw, 3)

moisture_clean <- moisture_raw %>%
  dplyr::select(
    datetime = `Measurement Time`,
    vwc      = `Measurement 1`,
    plot     = `Plot`,
    depth    = `depth`
  ) %>%
  filter(!is.na(vwc), !is.na(plot)) %>%
  mutate(
    plot  = as.numeric(plot),
    year  = year(datetime),
    depth = as.numeric(depth)
  ) %>%
  filter(year %in% c(2014, 2015))

# Seasonal means by plot, year, depth
moisture_summary <- moisture_clean %>%
  group_by(plot, year, depth) %>%
  summarise(
    vwc_mean = round(mean(vwc, na.rm = TRUE), 4),
    .groups  = 'drop'
  ) %>%
  pivot_wider(
    names_from   = depth,
    values_from  = vwc_mean,
    names_prefix = "vwc_"
  ) %>%
  rename(
    vwc_50cm  = vwc_50,
    vwc_100cm = vwc_100
  )

# Check
nrow(moisture_summary)
head(moisture_summary)
```
```{r}
moisture_summary <- moisture_summary %>%
  filter(plot %in% yield_clean$plot)

# Verify
nrow(moisture_summary)
distinct(moisture_summary, plot) %>% nrow()
```
```{r}
# Soils----

# Check 2014 soils
soils_2014_raw <- read_excel("data/raw/Soils.xlsx", 
                              sheet = "Soils-2014", skip = 12)
names(soils_2014_raw)
head(soils_2014_raw, 3)

soils_2014_clean <- soils_2014_raw %>%
  dplyr::select(
    plot       = `plot...1`,
    depth      = `depth`,
    NO3_N_ppm  = `nitrate...12`,
    NH4_N_ppm  = `ammonia`,
    LOI_OM_pct = `LOI OM`
  ) %>%
  filter(!is.na(plot), !is.na(depth)) %>%
  mutate(
    plot       = as.numeric(plot),
    depth      = str_trim(depth),
    NO3_N_ppm  = as.numeric(NO3_N_ppm),
    NH4_N_ppm  = as.numeric(NH4_N_ppm),
    LOI_OM_pct = as.numeric(LOI_OM_pct),
    year       = 2014
  ) %>%
  pivot_wider(
    names_from  = depth,
    values_from = c(NO3_N_ppm, NH4_N_ppm, LOI_OM_pct),
    names_sep   = "_"
  ) %>%
  rename_with(~ str_replace_all(., c(
    '_0-6"'   = '_0_6',
    '_6-18"'  = '_6_18',
    '_18-30"' = '_18_30'
  )))

# Check...The LOI_OM_pct deeper depths showing 32 NAs needs a note explaining the 2014 Agvise lab only measured OM at 0-6"
nrow(soils_2014_clean)
names(soils_2014_clean)

soils_2015_raw <- read_excel("data/raw/Soils.xlsx", 
                              sheet = "Soils-2015", skip = 10)
names(soils_2015_raw)
head(soils_2015_raw, 5)

soils_2015_clean <- soils_2015_raw %>%
  dplyr::select(
    sample_id  = `Sample`,
    LOI_OM_pct = `LOI OM`,
    NH4_N_ppm  = `NH4-N`,
    NO3_N_ppm  = `NO3-N`,
    TOC_pct    = `TOC`
  ) %>%
  # Remove header and NA rows
  filter(!is.na(sample_id), 
         sample_id != "ID",
         str_detect(sample_id, "\\d")) %>%
  mutate(
    # Extract plot number and depth code from sample ID
    plot       = as.numeric(str_extract(sample_id, "^\\d+")),
    depth_code = str_trim(str_extract(sample_id, "[ABC]$")),
    depth      = recode(depth_code,
                        "A" = '0-6"',
                        "B" = '6-18"',
                        "C" = '18-30"'),
    year       = 2015,
    # Average duplicate readings separated by "/"
    NO3_N_ppm  = map_dbl(str_split(as.character(NO3_N_ppm), "/"),
                         ~mean(as.numeric(str_trim(.x)), na.rm = TRUE)),
    NH4_N_ppm  = map_dbl(str_split(as.character(NH4_N_ppm), "/"),
                         ~mean(as.numeric(str_trim(.x)), na.rm = TRUE)),
    LOI_OM_pct = map_dbl(str_split(as.character(LOI_OM_pct), "/"),
                         ~mean(as.numeric(str_trim(.x)), na.rm = TRUE)),
    TOC_pct    = map_dbl(str_split(as.character(TOC_pct), "/"),
                         ~mean(as.numeric(str_trim(.x)), na.rm = TRUE))
  ) %>%
  filter(!is.na(plot)) %>%
  # Keep only plots in our experimental design
  filter(plot %in% yield_clean$plot) %>%
  dplyr::select(plot, year, depth, NO3_N_ppm, NH4_N_ppm, LOI_OM_pct, TOC_pct) %>%
  pivot_wider(
    names_from  = depth,
    values_from = c(NO3_N_ppm, NH4_N_ppm, LOI_OM_pct, TOC_pct),
    names_sep   = "_"
  ) %>%
  rename_with(~ str_replace_all(., c(
    '_0-6"'   = '_0_6',
    '_6-18"'  = '_6_18',
    '_18-30"' = '_18_30'
  )))

# Check
nrow(soils_2015_clean)
names(soils_2015_clean)

nrow(soils_2015_clean)

# Combine soils both years
soils_all <- bind_rows(soils_2014_clean, soils_2015_clean)
```

```{r}
# Join all datasets----
combined_clean <- yield_clean %>%
  left_join(lys_summary,      by = c("plot", "year")) %>%
  left_join(moisture_summary, by = c("plot", "year")) %>%
  left_join(soils_all,        by = c("plot", "year"))

# Final checks
nrow(combined_clean)
ncol(combined_clean)
names(combined_clean)

# Check missing values
combined_clean %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

# Save as the definitive final dataset
write.csv(combined_clean,
          "combined_clean_final.csv",
          row.names = FALSE)

cat("Final dataset saved\n")
cat("Rows:", nrow(combined_clean), "\n")
cat("Columns:", ncol(combined_clean), "\n")
```

