---
title: "Kernza Data Wrangle Project"
author: "Christopher SHerve"
date: "2026-04-10"
output: html_document
---

## Setup

```{r setup, include=FALSE}
# Packages to be installed 
library(tidyverse)
library(ggplot2)
library(googlesheets4)
library(lattice)
library(lubridate)
library(dplyr)

data <- read_sheet("https://docs.google.com/spreadsheets/d/1RZdL8GhR8bRt3f6G9pSMFs50jnjGvhdAajJhmgkZ0xs/edit?usp=sharing")
```
```{r}
# Make treatment a factor with readable labels----
data <- data %>%
  mutate(
    trt = factor(trt, 
                 levels = c(21, 22, 25, 23),
                 labels = c("0 kg N", "60 kg N", "Kernza/Alfalfa 40 kg N", "120 kg N")),
    year = factor(year),
    location = factor(location)
  )
str(data)
view(data)
## Data Overview
glimpse(data)
summary(data)
```


```{r}
# Long Format by adding Lysimete depths----
# ── STEP 1: Add n_fertilizer as numeric for regression ──────
# (it's already numeric but confirm it stayed that way after factoring)
str(data$n_fertilizer)  # should say num

# ── STEP 2: Create a monoculture-only subset for yield~N regression
data_mono <- data %>% filter(crops == "TLI")

# ── STEP 3: Create 2015-only subset for deeper soil depths
data_2015 <- data %>% filter(year == "2015")

# ── STEP 4: Create long format for soil depth comparisons
soil_long <- data %>%
  dplyr::select(year, location, plot, trt, crops, n_fertilizer,
                NO3_N_ppm_0_6, NO3_N_ppm_6_18, NO3_N_ppm_18_30) %>%
  pivot_longer(cols = c(NO3_N_ppm_0_6, NO3_N_ppm_6_18, NO3_N_ppm_18_30),
               names_to = "depth",
               values_to = "NO3_soil_ppm") %>%
  mutate(depth = recode(depth,
    "NO3_N_ppm_0_6"   = "0-6 inches",
    "NO3_N_ppm_6_18"  = "6-18 inches",
    "NO3_N_ppm_18_30" = "18-30 inches")) %>%
  filter(!is.na(NO3_soil_ppm))

# ── STEP 5: Quick check
glimpse(soil_long)

soil_long <- soil_long %>%
  mutate(depth = factor(depth, 
                        levels = c("0-6 inches", 
                                   "6-18 inches", 
                                   "18-30 inches")))
```
### Grain yield by treatment----
####Grain yield was compared across four treatments and two years using boxplots. Individual plot values are overlaid as points to show within-treatment variability. Year 3 (2014) and Year 4 (2015) are shown separately to capture stand age effects documented in Jungers et al. 2017.
```{r}
data %>%
  group_by(location, trt, year) %>%
  summarise(mean_lodging = round(mean(lodging_score, na.rm=TRUE), 2),
            mean_yield   = round(mean(grain_yield, na.rm=TRUE), 2))

# Summarize first
yield_summary <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    se_yield   = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

glimpse(yield_summary)
view(yield_summary)

# Plot
ggplot(yield_summary, aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8), 
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8),
                width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", 
                                "2015" = "#70AD47")) +
  labs(
    title   = "Mean Kernza grain yield by treatment, location, and year",
    subtitle = "Error bars show ± 1 standard error | n = 4 plots per treatment",
    x       = "Treatment",
    y       = "Mean grain yield (kg ha⁻¹)",
    fill    = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

```{r}
# Grain yield in relation to N-Trt, Location, Year, Lodging Score----
# Summarize both yield and lodging
yield_lodging_plot <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield   = mean(grain_yield, na.rm = TRUE),
    se_yield     = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    mean_lodging = mean(lodging_score, na.rm = TRUE),
    se_lodging   = sd(lodging_score, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Panel 1 — Yield
p1 <- ggplot(yield_lodging_plot, 
             aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6",
                                "2015" = "#70AD47")) +
  labs(title    = "Grain yield and lodging by treatment, location, and year",
       subtitle = "Waseca yield increased despite higher lodging — ruling out lodging as cause",
       x        = NULL,
       y        = "Mean yield (kg ha⁻¹)",
       fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

# Panel 2 — Lodging
p2 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_lodging, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_lodging - se_lodging,
                    ymax = mean_lodging + se_lodging),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6",
                                "2015" = "#70AD47")) +
  labs(x    = "Treatment",
       y    = "Mean lodging score",
       fill = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

# Stack them vertically using patchwork
install.packages("patchwork")
library(patchwork)
p1 / p2


combined_plot <- p1 / p2

ggsave(filename = "~/GitHub/kernza_data_wrangle/dw_kernza_project/yield_lodging_by_location_year.png",
       plot     = combined_plot,
       width    = 10,
       height   = 8,
       dpi      = 300)

```

### Soil moisture by depth and location
#### pivot_longer is the key here this data has two separate columns (vwc_50cm and vwc_100cm) but you want to plot them together. This reshapes the data so depth becomes one column and the moisture value becomes another. Then it plots both depths as lines over time, colored by location (Waseca vs Lamberton), with separate panels per depth.
```{r}
data %>%
  pivot_longer(cols = c(vwc_50cm, vwc_100cm),
               names_to = "depth", values_to = "vwc") %>%
  ggplot(aes(x = year, y = vwc, color = location, group = interaction(location, depth))) +
  geom_point() + geom_line() +
  facet_wrap(~depth) +
  labs(title = "Soil moisture by depth and location",
       y = "Volumetric water content") +
  theme_bw()
```

### Nitrate leaching by treatment and year
```{r}
ggplot(data, aes(x = trt, y = NO3_mean, fill = year)) +
  geom_boxplot() +
  labs(title = "Nitrate leaching by treatment",
       x = "Treatment", y = "NO3 (mg/L)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Biomass proportion by tissue type
```{r}
data %>%
  select(year, location, trt, grain_yield, iwg_straw_yield, legume_biomass) %>%
  pivot_longer(cols = c(grain_yield, iwg_straw_yield, legume_biomass),
               names_to = "tissue", values_to = "biomass") %>%
  ggplot(aes(x = trt, y = biomass, fill = tissue)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~year) +
  labs(title = "Proportion of biomass by tissue type",
       y = "Proportion", x = "Treatment") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#Have to be opened highlighted together
library(emmeans)
library(multcomp)
library(multcompView)
```

## ANOVA Research Questions

### Question 1: Does grain yield differ by treatment?

**Hypothesis:** Grain yield varies significantly among nitrogen fertilizer treatments.

```{r}
model1 <- aov(grain_yield ~ trt, data = data)
summary(model1)

em1 <- emmeans(model1, ~ trt)
cld(em1, Letters = letters)
```

### Question 2: Does nitrate leaching differ by treatment?

**Hypothesis:** Nitrate leaching (NO3) varies by treatment due to differences in N input.

```{r}
model2 <- aov(NO3_mean ~ trt, data = data)
summary(model2)

em2 <- emmeans(model2, ~ trt)
cld(em2, Letters = letters)
```

### Question 3: Does plant height differ by location?

**Hypothesis:** Plant height differs between Waseca and Lamberton due to site conditions.

```{r}
model3 <- aov(plant_height ~ location, data = data)
summary(model3)

em3 <- emmeans(model3, ~ location)
cld(em3, Letters = letters)
```

## Regression Research Questions

### Question 1: Does grain yield increase with N fertilizer rate?

**Hypothesis:** Grain yield increases with nitrogen fertilizer rate.

```{r}
reg1 <- lm(grain_yield ~ n_fertilizer, data = data)
summary(reg1)

ggplot(data, aes(x = n_fertilizer, y = grain_yield)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Grain yield vs N fertilizer rate",
       x = "N fertilizer (kg/ha)", y = "Grain yield (kg/ha)") +
  theme_bw()
```

### Question 2: Does soil moisture affect nitrate leaching?

**Hypothesis:** Higher soil moisture is associated with greater nitrate leaching.

```{r}
reg2 <- lm(NO3_mean ~ vwc_100cm, data = data)
summary(reg2)

ggplot(data, aes(x = vwc_100cm, y = NO3_mean)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Nitrate leaching vs soil moisture at 100cm",
       x = "VWC at 100cm", y = "NO3 mean (mg/L)") +
  theme_bw()
```

### Question 3: Does lodging score increase with plant height?

**Hypothesis:** Taller plants are more prone to lodging.

```{r}
reg3 <- lm(lodging_score ~ plant_height, data = data)
summary(reg3)

ggplot(data, aes(x = plant_height, y = lodging_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Lodging score vs plant height",
       x = "Plant height (cm)", y = "Lodging score") +
  theme_bw()
```

## Data overview

```{r}
glimpse(data)
```
Then click the Knit button at the top (blue yarn icon) to make sure it runs.
Go ahead and create that file — what packages were you planning to use? Share that list and I'll write out your full starter template!