---
title: "Kernza Data Wrangle Project Analysis"
author: "Christopher Sherve"
date: "`r Sys.Date()`"
output: html_document
---

## Setup

Packages are loaded and data is read from Google Sheets. The `emmeans`, 
`multcomp`, and `multcompView` packages are loaded later immediately before 
the ANOVA section to avoid the MASS package masking `dplyr::select()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(googlesheets4)
library(lubridate)
library(patchwork)
library(knitr)
library(lattice)

# Resolve MASS package conflict with dplyr
select <- dplyr::select
filter <- dplyr::filter

# Load corrected dataset from Google Sheets
data <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1RZdL8GhR8bRt3f6G9pSMFs50jnjGvhdAajJhmgkZ0xs/edit?usp=sharing",
  sheet = "combined_clean_final_v2"
)
```

```{r factor-variables}
data <- data %>%
  mutate(
    trt = factor(trt,
                 levels = c(21, 22, 25, 23),
                 labels = c("0 kg N", "60 kg N",
                            "Kernza/Alfalfa 40 kg N", "120 kg N")),
    year     = factor(year),
    location = factor(location),
    environment = paste(location, year, sep = "_")
  )
```

```{r data-subsets, include=FALSE}
data_mono <- data %>% filter(crops == "TLI")
data_2015 <- data %>% filter(year == "2015")
```

---

## Data Overview

The primary dataset (`combined_clean_final_v2`) integrates yield, lysimeter 
nitrate, soil moisture, and soil chemistry data from Waseca and Lamberton, MN 
across 2014 and 2015 (Years 3 and 4 of the study). Full data cleaning and 
QA/QC steps are documented in `data_wrangling.Rmd`, `metadata_kernza.md`, 
and `lysimeter_QAQC_findings.md`.

```{r data-overview}
glimpse(data)
summary(data)
```

---

## Data Limitations

The following limitations were identified during data wrangling and QA/QC 
and shape the interpretation of the analyses that follow:

1. **Soil organic matter at depth** Loss-on-ignition organic matter is 
   only available at 0-6 inch depth for 2014 (Agvise Laboratories). Deeper 
   depths (6-18" and 18-30") are available for 2015 only

2. **Soil moisture in control plots** The was no data collected for the control 
   treatment plots (trt 21, 0 kg N) by design. VWC data unavailable 
   for 16 plots, preventing full assessment of moisture × N rate interactions.

3. **Lysimeter data interpretation** Lysimeters measure NO3 concentration 
   in soil water at 100cm depth. August 2014 is absent due to harvest logistics. 
   For these reasons formal statistical analysis of lysimeter data is not 
   performed.

4. **Nitrogen budget in intercrop plots** The intercrop treatment receives 
   40 kg N ha⁻¹ of synthetic fertilizer plus variable additional nitrogen 
   from alfalfa biological fixation. Timing of N availability cannot be 
   quantified.

---

## Exploratory Visualizations

### Grain Yield by Treatment, Location, and Year

Grain yield was compared across four treatments and two years. Waseca and 
Lamberton are shown as separate panels to reveal the site × year interaction. 
Error bars show ± 1 standard error (n = 4 plots per treatment per location).

```{r yield-bar-plot, fig.width=9, fig.height=5}
yield_summary <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    se_yield   = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

ggplot(yield_summary, aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Mean Kernza grain yield by treatment, location, and year",
    subtitle = "Error bars show ± 1 SE | n = 4 plots per treatment",
    x        = "Treatment",
    y        = "Mean grain yield (kg ha⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

Lamberton yields declined dramatically (which is expected) from 2014 to 2015 across all treatments 
while Waseca yields increased. This opposite directional response between sites 
demonstrates a strong treatment × environment interaction and not expected. Site-level 
differences in soil drainage and seasonal precipitation distribution could 
drive the divergent yield responses.

---

### Grain Yield and Lodging by Treatment, Location, and Year

To evaluate whether lodging contributed to yield differences between years 
mean grain yield and lodging score are plotted together. If lodging caused 
yield decline, sites with higher lodging should show lower yield. Waseca shows 
the opposite pattern where yield increased from 2014 to 2015 despite increasing 
lodging scores and ruling out lodging as the primary cause of the Waseca yield 
improvement.

```{r yield-lodging-plot, fig.width=10, fig.height=8}
yield_lodging_plot <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield   = mean(grain_yield, na.rm = TRUE),
    se_yield     = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    mean_lodging = mean(lodging_score, na.rm = TRUE),
    se_lodging   = sd(lodging_score, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

p1 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Grain yield and lodging by treatment, location, and year",
       subtitle = "Waseca yield increased despite higher lodging — lodging not the cause",
       x        = NULL,
       y        = "Mean yield (kg ha⁻¹)",
       fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

p2 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_lodging, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_lodging - se_lodging,
                    ymax = mean_lodging + se_lodging),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(x    = "Treatment",
       y    = "Mean lodging score (0-10)",
       fill = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

p1 / p2
```

---

### Soil Water NO3 at 100cm Depth Exploratory Only

Lysimeter NO3 concentrations (mg/L) from soil water samples at 100cm depth 
are shown below as exploratory context. Statistical analysis is not 
performed on this data due to the limitations noted above (single depth, 
concentration, seasonal averaging, August 2014 gap, and high 
below detection limit (BDL) frequency. Patterns suggest site differences between Waseca and 
Lamberton but should be interpreted cautiously.

```{r no3-plot, fig.width=9, fig.height=5}
ggplot(data, aes(x = trt, y = NO3_mean_mgL, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Soil water NO3 at 100cm depth by treatment, location, and year",
    subtitle = "Exploratory visualization only — not used for formal inference",
    x        = "Treatment",
    y        = "Mean NO3 concentration (mg L⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

---

### Soil Moisture Field Map

Seasonal mean volumetric water content (VWC) is displayed as a spatial 
field map by plot, treatment, depth, location, and year. Each cell 
represents one plot colored by its seasonal mean moisture. Grey cells 
labeled "NS" indicate no sensor data was available in the 
0 kg N control treatment. There seems to be a bit of spatial variation which is exhibited by the blocking. 

```{r moisture-field-map, fig.width=12, fig.height=8}
moisture_field_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, vwc_100cm) %>%
  pivot_longer(cols = c(vwc_50cm, vwc_100cm),
               names_to  = "depth",
               values_to = "vwc") %>%
  mutate(
    depth = recode(depth,
                   "vwc_50cm"  = "50cm depth",
                   "vwc_100cm" = "100cm depth"),
    depth = factor(depth, levels = c("50cm depth", "100cm depth")),
    block = paste("Block", replicate)
  ) %>%
  group_by(location, year, depth, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup()

ggplot(moisture_field_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)), 
           fill = vwc)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(plot, "\n", 
                                ifelse(is.na(vwc), "NS", 
                                       round(vwc, 3)))),
            size = 2.3, lineheight = 0.9) +
  facet_grid(location ~ year + depth) +
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    values   = scales::rescale(c(0.13, 0.18, 0.22, 0.26, 0.30, 0.37)),
    limits   = c(0.13, 0.37),
    na.value = "grey80",
    name     = "VWC\n(m³/m³)",
    breaks   = c(0.15, 0.20, 0.25, 0.30, 0.35)) +
  labs(
    title    = "Soil moisture field map by plot location",
    subtitle = "Each cell = one plot | Grey cells (NS) = control plots with no sensors",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 9),
    axis.text        = element_text(size = 8),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    aspect.ratio     = 0.7)
```

---

### Spatial Relationship Moisture, Yield, and Treatment

This three-layer visualization combines soil moisture (tile color), grain 
yield (circle size), and treatment (circle color) in the same spatial 
field layout. This allows simultaneous examination of moisture patterns, 
yield variation, and treatment effects at the plot level.

```{r moisture-yield-trt-map, fig.width=14, fig.height=9, dpi=150, out.width="100%"}
moisture_yield_trt_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, grain_yield) %>%
  mutate(block = paste("Block", replicate)) %>%
  group_by(location, year, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup() %>%
  filter(!is.na(grain_yield))

ggplot(moisture_yield_trt_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)))) +
  geom_tile(aes(fill = vwc_50cm), color = "white", linewidth = 1.5) +
  geom_point(aes(size = grain_yield, color = trt), 
             alpha = 0.85, stroke = 1.2) +
  geom_text(aes(label = plot), 
            size = 2, color = "black",
            fontface = "bold", vjust = 2.8) +
  facet_grid(location ~ year) +
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    limits   = c(0.13, 0.37),
    na.value = "grey85",
    name     = "VWC 50cm\n(m³/m³)") +
  scale_color_manual(values = c(
    "0 kg N"                 = "#2C3E50",
    "60 kg N"                = "#3498DB",
    "Kernza/Alfalfa 40 kg N" = "#27AE60",
    "120 kg N"               = "#C0392B"),
    name = "Treatment") +
  scale_size_continuous(
    range = c(3, 12),
    name  = "Grain yield\n(kg ha⁻¹)",
    breaks = c(50, 150, 300, 500)) +
  labs(
    title    = "Spatial relationship: soil moisture, yield, and treatment",
    subtitle = "Tile color = soil moisture | Circle size = yield | Circle color = treatment",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    legend.box       = "vertical") +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    color = guide_legend(order = 3, override.aes = list(size = 5)))
```

---

---

## ANOVA Research Questions

Packages for post-hoc comparisons are loaded here to avoid the MASS package 
masking `dplyr::select()` earlier in the document.

```{r load-anova-packages, message=FALSE, warning=FALSE}
library(emmeans)
library(multcomp)
library(multcompView)
```

---

### ANOVA Question 1: Does treatment affect grain yield and does this interact with environment?

**Hypothesis:** The four treatments (0 kg N, 60 kg N, 120 kg N, and 
Kernza/Alfalfa 40 kg N intercrop) will produce different grain yields, 
and this effect will vary by environment (location × year). Specifically:
- N fertilizer rate is expected to increase yield up to some optimized point
- The intercrop treatment is hypothesized to perform similarly or better than 60 kg N and 120 kg N
  through N fixation by alfalfa.
- Optimizing nitrogen treatment for IWG is also being tested.  
- The intercrop advantage may diminish by year 4 (2015) as competition 
  with the aging Kernza stand increases

```{r anova-q1, fig.width=10, fig.height=5}
model1 <- aov(grain_yield ~ trt * location * year, data = data)
summary(model1)

em1 <- emmeans(model1, ~ trt | location * year)
cld(em1, Letters = letters)

data %>%
  ggplot(aes(x = factor(year), y = grain_yield, fill = trt)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.85)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c(
    "0 kg N"                 = "#2C3E50",
    "60 kg N"                = "#3498DB",
    "Kernza/Alfalfa 40 kg N" = "#2ECC71",
    "120 kg N"               = "#C0392B")) +
  labs(title    = "Grain yield by treatment, location, and year",
       subtitle = "All four treatments shown | Intercrop = Kernza/Alfalfa 40 kg N",
       x        = "Year",
       y        = "Grain yield (kg ha⁻¹)",
       fill     = "Treatment") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"),
        legend.position  = "right")
```

**Interpretation:** The ANOVA revealed two highly significant effects. 
The main effect of treatment (p = 0.0037) and the location × year 
interaction (p = 0.0003). The significant treatment effect confirms 
that the four treatments produced meaningfully different grain yields 
across the dataset. The location × year interaction reflects the opposite 
directional yield response between sites. Waseca yields increased from 
2014 to 2015 while Lamberton yields decreased. The absence of a 
significant treatment × location or treatment × year interaction 
(p = 0.82 and p = 0.70 respectively) indicates that treatment effects 
were relatively consistent across environments. 

The boxplot reveals several important patterns: the 0 kg N control 
consistently produced the lowest yields in all environments. N fertilizer 
rate generally increased yield with some diminishing returns at 120 kg N.  
The intercrop treatment (40 kg N synthetic + alfalfa N fixation) 
matched or exceeded the 120 kg N monoculture at Lamberton 2014, 
suggesting substantial alfalfa N contribution in the year 3 environment 
with highest yield potential. 

The intercropped Kernza treatment does have far less variation in terms of yield
in comparison to the non intercropped treatments.  Given the flooding occured in 
Waseca, it is safe to say based on this data it would be beneficial to have an
intercropped treatment that matches a mono culture treatment.  Also having an 
intercropped treatment with no nitrogen application.  

**Supplementary 2-level analysis crop system effect:** For comparison, 
an additional ANOVA collapses the three N rate monoculture treatments 
(0, 60, 120 kg N) into a single "TLI monoculture" group to test whether 
the presence of a legume intercrop alone affects yield, independent of 
N rate variation.

```{r anova-q1-crops, fig.width=8, fig.height=5}
model1_crops <- aov(grain_yield ~ crops * location * year, data = data)
summary(model1_crops)

em1_crops <- emmeans(model1_crops, ~ crops | location * year)
cld(em1_crops, Letters = letters)

data %>%
  ggplot(aes(x = factor(year), y = grain_yield, fill = crops)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("TLI" = "#2E75B6",
                                "TLI/leg" = "#2ECC71")) +
  labs(title    = "Grain yield by crop system across environments",
       subtitle = "TLI = Kernza monoculture (pooled N rates) | TLI/leg = Kernza/alfalfa intercrop",
       x        = "Year",
       y        = "Grain yield (kg ha⁻¹)",
       fill     = "Crop system") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"))
```

Note that the crop system model shows a marginal effect (p = 0.073) 
compared to the highly significant full treatment model (p = 0.0037). 
The reduced statistical power reflects the fact that pooling the three 
N rates introduces substantial within-group variance, obscuring the 
clearer treatment-level effects visible in the 4-treatment analysis.

---

### ANOVA Question 2: Does plant height differ by treatment?

**Hypothesis:** Plant height will increase with nitrogen fertilizer rate 
as higher N availability promotes vegetative growth. The intercrop treatment 
may show different height dynamics due to competition with alfalfa.

```{r anova-q2, fig.width=8, fig.height=5}
model2 <- aov(plant_height ~ trt * year*location, data = data)
summary(model2)

em2 <- emmeans(model2, ~ trt*year*location)
cld(em2, Letters = letters)

ggplot(data, aes(x = trt, y = plant_height, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Plant height by treatment and year",
       subtitle = "Height measured from soil to top of seed head at physiological maturity",
       x        = "Treatment",
       y        = "Plant height (cm)",
       fill     = "Year") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```
**Interpretation:** it is significant that treatment of nitrogen application will 
improve plant height versus no nitrogen application.  There is not a significant 
difference between the nitrogen applied treatments, but the intercropped treatment 
does have a higher averaged plant height than the monocropped treatments.  
I think the obvious question goes back to having a intercropped treatmnet that matches 
one of the monocropped treatments and comparing yield and plant height.  

### Part 2 of ANOVA Question 2: Does plant height differ by treatment?
How well does plant height corrleate to yield in year 1 to year 2 
```{r}
xyplot(grain_yield ~ plant_height | location * year, 
            data = data, 
            type = c("p", "r"))
```
**Interpretation:** so while there is correlation between grain yield and plant 
height, the telling take away is the difference between locations in 2015.  
IWG behaves normally in Lamberton with good yield to plant height ration in 2014.  
The flood of 2014 in Waseca had a perfound effect on physiological status of the 
IWG plants because in 2015 the yield is opposite to normal year two yield behavior 
in IWG which is shown in Lamberton in 2015 where plant height stayed the same and 
yield decreased.  This is important to note because the nutrient use is being limited 
by something and it is not plant height in this case.  
---

## Regression Research Questions

### Regression Question 1: Does grain yield increase with N fertilizer rate?

**Hypothesis:** Grain yield increases linearly with N fertilizer rate in 
Kernza monoculture plots. Based on Jungers et al. (2017) a quadratic 
relationship is also possible if yields decline at high N rates due to 
lodging. The linear model is tested first.

```{r regression-q1, fig.width=8, fig.height=5}
reg1 <- lm(grain_yield ~ n_fertilizer+location, data = data_mono)

summary(reg1)

ggplot(data_mono, aes(x = n_fertilizer, y = grain_yield)) +
  geom_point(aes(color = location, shape = year), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Kernza grain yield vs N fertilizer rate (monoculture only)",
    subtitle = "Linear regression | Points colored by location, shaped by year",
    x        = "N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year") +
  theme_bw()
```

**Supplementary regression (Lamberton only):** Because Waseca 2014 
experienced documented historic flooding (Minnesota DNR 2014; 12.31 inches 
of rainfall in June, the wettest month in Minnesota state history) that 
suppressed reproductive development independently of N rate, a Lamberton-only 
supplementary regression provides a cleaner test of typical Kernza N response 
under non-extreme conditions. Year is included as a covariate to capture 
the documented year 3-to-4 stand age effect.

```{r regression-q1-lamberton, fig.width=8, fig.height=5}
reg1_lamberton <- lm(grain_yield ~ n_fertilizer + year, 
                     data = data_mono %>% filter(location == "Lamberton"))
summary(reg1_lamberton)

ggplot(data_mono %>% filter(location == "Lamberton"), 
       aes(x = n_fertilizer, y = grain_yield, color = year)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Lamberton-only: Kernza grain yield vs N fertilizer rate",
    subtitle = "Separate regression lines for 2014 (year 3) and 2015 (year 4)",
    x        = "N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)",
    color    = "Year") +
  theme_bw()
```

**Interpretation:** The Lamberton only regression reveals two key patterns. 
The N rate effect is positive but marginal (slope = 0.98 kg yield per kg N, 
p = 0.098), indicating each additional kg of N fertilizer is associated 
with approximately 1 kg/ha more grain yield. The year effect is highly 
significant (-143 kg/ha in 2015 compared to 2014, p = 0.017), confirming 
the documented year 3-to-4 Kernza yield decline. This suggests that for 
established Kernza stands at this age, fertilization cannot fully compensate for 
the natural year-over-year yield decline. The marginal significance of the N rate 
slope reflects the limited statistical power of the Lamberton-only 
subset (n = 23 observations).  Additionally, there is evidence of a optimal 
around 60-80 kg N ha⁻¹ but the variation is so wide it's very difficult to 
determine an optimized amount without further testing. 

---

### Regression Question 2: Where does the intercrop treatment fall on the yield ~ N rate curve?

**Hypothesis:** If the alfalfa intercrop provides a yield benefit through 
N fixation, the intercrop treatment (40 kg N ha⁻¹ synthetic + alfalfa-fixed N) 
should produce yields comparable to or exceeding monoculture plots receiving 
similar or higher synthetic N rates. If competition effects dominate between IWG and alfalfa,
intercrop yields will fall below the monoculture regression line at 40 kg N ha⁻¹.

```{r regression-q2, fig.width=8, fig.height=5}
reg2 <- lm(grain_yield ~ n_fertilizer*year*location, data = data_mono)
summary(reg2)
ggplot() +
  # This matches your lm(grain_yield ~ n_fertilizer * year * location)
  geom_smooth(data = data_mono,
              aes(x = n_fertilizer, y = grain_yield, color = location, linetype = year),
              method = "lm", se = TRUE, alpha = 0.1) +
  
  # Monoculture points split by location and year to show the "spread"
  geom_point(data = data_mono,
             aes(x = n_fertilizer, y = grain_yield, color = location, shape = year),
             size = 2, alpha = 0.5) +
  
  # The Intercrop points (Green Triangles) sitting at 40kg N
  geom_point(data = data %>% filter(crops == "TLI/leg"),
             aes(x = n_fertilizer, y = grain_yield),
             color = "#2ECC71", size = 4, shape = 17) +
  
  # Custom colors to match your previous theme
  scale_color_manual(values = c("Waseca" = "#2E75B6", "Lamberton" = "#E74C3C")) +
  
  labs(
    title    = "Intercrop Yield vs. Complex Monoculture N-Response",
    subtitle = "3-way interaction model coefficients",
    x        = "N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)",
    color    = "Location",
    linetype = "Year",
    shape    = "Year") +
  theme_bw()

```
**Interpretation:** The three way interaction model reveals that the N-response 
slope for Waseca 2015 (Blue Dashed) was steeper than for Lamberton 2015 
(Red Dashed), supporting the theory that the 2014 flood 'reset' the physiological 
budget. Furthermore, the intercrop treatment (Green Triangles) at 40 kg N 
consistently outperformed the predicted monoculture yield across most environments, 
demonstrating that biological N-fixation may be more efficient than synthetic N 
in aging perennial stands
---

### Regression Question 3: Does lodging score increase with plant height?

**Hypothesis:** Taller plants are more prone to lodging due to a higher 
center of gravity and increased wind resistance. However Jungers et al. 
(2017) found that plant height did not explain additional variation in 
lodging beyond the effect of N fertilizer rate, suggesting stem diameter 
or composition may be more important than height alone.

```{r regression-q3, fig.width=8, fig.height=5}
reg3 <- lm(lodging_score ~ plant_height, data = data)
summary(reg3)

ggplot(data, aes(x = plant_height, y = lodging_score)) +
  geom_point(aes(color = trt), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = c(
    "0 kg N"                 = "#D3D3D3",
    "60 kg N"                = "#74A9CF",
    "Kernza/Alfalfa 40 kg N" = "#2ECC71",
    "120 kg N"               = "#2C7BB6")) +
  labs(
    title    = "Lodging score vs plant height",
    subtitle = "Does plant height predict lodging susceptibility?",
    x        = "Plant height (cm)",
    y        = "Lodging score (0-10)",
    color    = "Treatment") +
  theme_bw()
```
**Interpretation:**The lodging resilience of the Kernza/Alfalfa intercrop is 
likely a synergistic result of structural facilitation (physical support from 
alfalfa stems) and nutritional moderation. Unlike the 120 kg N treatment, which 
promotes rapid, structurally weak cell elongation, the intercrop system provides
a slow release nitrogen source that may favor thicker cell walls and better lignification
---

### Regression Question 4: Does alfalfa legume biomass correlate with Kernza straw yield in intercrop plots?

**Hypothesis:** Straw yield is examined in this analysis (rather than grain 
yield) because straw represents the vegetative biomass response to N 
availability. If alfalfa N fixation is the dominant intercrop benefit, 
greater alfalfa biomass should correlate with greater Kernza straw yield 
because more available N drives vegetative growth. Alternatively, if 
resource competition dominates the intercrop relationship, greater alfalfa 
biomass may correlate with reduced Kernza straw yield. Examining straw 
rather than grain isolates the vegetative response from reproductive 
allocation effects, which are addressed separately in the supplementary 
straw-to-grain conversion analysis below.

**Statistical limitations:** This analysis uses only the 16 intercrop 
plots (n = 4 per environment × 4 environments). Statistical power is 
limited, and the location and year effects are expected to dominate 
the raw correlation. The regression below controls for location and year.

```{r regression-q4, fig.width=8, fig.height=5}
data_intercrop <- data %>% 
  filter(crops == "TLI/leg", !is.na(legume_biomass), !is.na(grain_yield))

reg4 <- lm(iwg_straw_yield~legume_biomass*year*location, data=data_intercrop)
summary(reg4)



ggplot(data_intercrop, 
       aes(x = legume_biomass, y = iwg_straw_yield, 
           color = location, shape = factor(year))) +
  geom_point(size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, aes(group = location)) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Kernza straw yield vs alfalfa biomass in intercrop plots",
    x        = "Alfalfa biomass (kg ha⁻¹)",
    y        = "Kernza straw yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year") +
  theme_bw()
```
```{r}
# Checking the grain yield to straw ratio to confirm the year two reduction
reg5 <- lm(grain_yield~iwg_straw_yield*location*year, data=data)
summary(reg5)

ggplot(data, aes(x = iwg_straw_yield, y = grain_yield, color = location)) +
  # This matches your lm(grain_yield ~ iwg_straw_yield * location * year)
  geom_smooth(method = "lm", aes(linetype = year), se = TRUE, alpha = 0.1, size = 1.2) +
  
  # Individual plot observations
  geom_point(aes(shape = year), size = 3, alpha = 0.6) +
  
  # Custom colors to stay consistent with your theme
  scale_color_manual(values = c("Waseca" = "#2E75B6", "Lamberton" = "#E74C3C")) +
  
  labs(
    title    = "The Conversion Efficiency: Grain Yield vs. Straw Yield",
    subtitle = "Steeper slopes = Higher ROI | Flatter slopes = High 'Root Tax' competition",
    x        = "IWG Straw Yield (kg ha⁻¹)",
    y        = "IWG Grain Yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year",
    linetype = "Year") +
  theme_bw()
```


**Interpretation:** The conversion efficiency of straw-to-grain reveals two 
distinct biological pathways. At Lamberton, the transition from 2014 to 2015 
(Dashed Red) shows a significant flattening of the efficiency slope, representing 
the high 'Root Tax' of an aging stand. In contrast, Waseca 2015 (Dashed Blue) 
shows a steeper, more efficient rebound, supporting the 'Flood Reset' hypothesis 
where early stand stress temporarily preserved the plant's reproductive efficiency."



---

## Summary of Findings

```{r summary-table}
tibble(
  Question = c(
    "ANOVA Q1: Treatment × environment effect on grain yield",
    "ANOVA Q2: Treatment effect on plant height (with grain yield correlation)",
    "Regression Q1: Grain yield ~ N fertilizer rate (monoculture, with Lamberton-only supplementary)",
    "Regression Q2: Intercrop position on yield curve",
    "Regression Q3: Lodging score ~ plant height",
    "Regression Q4: Kernza straw yield ~ alfalfa legume biomass (intercrop plots, with grain:straw efficiency)"
  ),
  `Response Variable` = c(
    "grain_yield", "plant_height",
    "grain_yield", "grain_yield", "lodging_score", "iwg_straw_yield / grain_yield"
  )
) %>%
  kable(caption = "Summary of research questions and response variables")
```

---

## Future Directions

### Context: Documented June 2014 flooding at Waseca

The opposite yield trajectories between Waseca (2014→2015: +47% grain yield) 
and Lamberton (2014→2015: -49% to -66% grain yield across treatments) are 
the most striking pattern in this dataset. Weather records clarify a key 
driver at Waseca: June 2014 was the wettest month in Minnesota state 
history, with Waseca receiving 12.31 inches of rainfall is approximately 
four times the historical June average (Minnesota DNR 2014). Multiple flash 
flooding events affected Waseca County during the month, with flooding 
documented by National Weather Service alerts. The flooding occurred 
during the critical stem elongation and early reproductive development 
stages of Kernza on the poorly drained Webster clay loam soil, producing 
prolonged saturated and anaerobic conditions. The 2014 Waseca yields 
therefore represent suppressed yields under a documented extreme flooding 
event rather than typical year 3 Kernza performance.

The observation that 2015 Waseca grain yields nearly doubled while plant 
height, straw biomass, and lodging remained relatively stable suggests 
the crop expressed more of its reproductive potential once flooding stress 
was removed. With Lamberton's contrasting 2014-2015 decline under normal 
conditions raises testable questions about the mechanistic basis of 
year-over-year yield variation in perennial Kernza stands.

### Research priority 1: Does root biomass accumulation drive the year 
1-to-year 2 Kernza yield decline?

The well-documented year 1-to-year 2 grain yield decline in Kernza stands 
remains mechanistically unresolved and is identified as a key challenge 
for perennial grain crop development (Cattani & Asselin 2018; De Pauw 
et al. 2022). One plausible hypothesis emerges from established principles 
of perennial grass physiology.

Root maintenance respiration in perennial plants can consume over 50% of 
the carbon utilized within the root system (Eissenstat 1993). Because this 
maintenance cost scales with absolute root biomass, a larger year 2 Kernza 
root system would impose a substantially higher absolute carbon demand than 
a smaller year 1 system, even if proportional allocation rules remained 
unchanged. Similarly, nitrogen remobilization from vegetative tissues to 
developing grain (accounting for 65-90% of grain N in cereals; Hu et al. 
2016) must compete with N demands from an expanding root system.

Under this source-sink competition framework, the year 1-to-2 yield 
decline may result from increased absolute carbon and nitrogen consumption 
by root maintenance rather than from changes in allocation priorities. 
The plant's photosynthetic capacity remains similar, but a larger fraction 
of fixed photosynthate and cycled nitrogen is consumed by root maintenance 
before it becomes available for grain filling.

**Testable predictions:**

- Root biomass should increase substantially between year 1 and year 2 
  Kernza stands (already documented by Jungers and Gutknecht)
- Absolute carbon allocation to root maintenance respiration should 
  increase proportionally with root biomass
- Grain yield should correlate inversely with root maintenance carbon 
  demand across stand ages
- Nitrogen remobilization efficiency to grain should decrease as root 
  system N sink strength increases

**Methods:** Destructive root biomass sampling, root respiration 
measurements (chamber or isotope-tracer methods), whole-plant carbon 
allocation studies using 13C labeling, and plant tissue N analysis 
at key phenological stages. These methods are established in related 
perennial grass research (Carbone et al. 2007; Yao and Chen 2019) and 
transferable to Kernza.

**Status in literature:** No published study has directly tested this 
relationship in Kernza. Existing Kernza research documents root biomass 
increases with stand age but frames this as beneficial for soil carbon 
storage rather than as a potential yield-limiting cost. This represents 
a specific, tractable research gap.

### Research priority 2: What regulatory pathways control root biomass 
development in Kernza, and how does waterlogging stress alter them?

This priority is contingent on Priority 1 findings. If root biomass 
accumulation is confirmed as a significant contributor to yield decline, 
identifying regulatory pathways that control root biomass development 
could inform breeding targets for improved yield stability across years.

The 2014 Waseca flooding provides a natural experimental context. The 
flooding reduced grain yield while apparently limiting normal root 
expansion. Understanding the molecular basis of this response has 
implications for both site selection (Kernza is promoted for marginal 
sites that may include poorly drained areas) and breeding for 
waterlogging tolerance.

**Approach:** Controlled waterlogging experiments using established 
methods (container flooding, hydroponic hypoxia, rhizotrons; Barickman 
et al. 2019; Kong et al. 2022) could recreate the 2014 Waseca conditions 
in a controlled setting. Comparison of root growth, development, and 
architecture between stressed and unstressed plants would provide the 
phenotypic baseline. If phenotypic differences are substantial, targeted 
gene expression analysis could investigate candidate pathways including 
hypoxia response (ERF-VII transcription factors), root meristem 
regulation, and carbohydrate allocation signaling.

**Research sequence rationale:** Priority 1 must precede Priority 2 
because phenotypic characterization of the root-yield relationship 
provides the foundation for any molecular investigation. Investing in 
gene expression work before confirming the phenotypic relationship risks 
targeting the wrong mechanism.
---

*Data wrangling and QA/QC fully documented in `data_wrangling.Rmd`.*  
*Reference: Jungers et al. 2017. Agron. J. 109:462–472.*
