---
title: "Kernza Data Wrangle Project — Analysis"
author: "Christopher Sherve"
date: "`r Sys.Date()`"
output: html_document
---

## Setup

Packages are loaded and data is read from Google Sheets. The `emmeans`, 
`multcomp`, and `multcompView` packages are loaded later immediately before 
the ANOVA section to avoid the MASS package masking `dplyr::select()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(googlesheets4)
library(lubridate)
library(patchwork)
library(knitr)

# Resolve MASS package conflict with dplyr
select <- dplyr::select
filter <- dplyr::filter

# Load corrected dataset from Google Sheets
data <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1RZdL8GhR8bRt3f6G9pSMFs50jnjGvhdAajJhmgkZ0xs/edit?usp=sharing",
  sheet = "combined_clean_final_v2"
)
```

```{r factor-variables}
data <- data %>%
  mutate(
    trt = factor(trt,
                 levels = c(21, 22, 25, 23),
                 labels = c("0 kg N", "60 kg N",
                            "Kernza/Alfalfa 40 kg N", "120 kg N")),
    year     = factor(year),
    location = factor(location),
    environment = paste(location, year, sep = "_")
  )
```

```{r data-subsets, include=FALSE}
data_mono <- data %>% filter(crops == "TLI")
data_2015 <- data %>% filter(year == "2015")
```

---

## Data Overview

The primary dataset (`combined_clean_final_v2`) integrates yield, lysimeter 
nitrate, soil moisture, and soil chemistry data from Waseca and Lamberton, MN 
across 2014 and 2015 (Years 3 and 4 of the study). Full data cleaning and 
QA/QC steps are documented in `data_wrangling.Rmd`, `metadata_kernza.md`, 
and `lysimeter_QAQC_findings.md`.

```{r data-overview}
glimpse(data)
summary(data)
```

---

## Data Limitations

The following limitations were identified during data wrangling and QA/QC 
and shape the interpretation of the analyses that follow:

1. **Soil organic matter at depth** — Loss-on-ignition organic matter is 
   only available at 0-6 inch depth for 2014 (Agvise Laboratories). Deeper 
   depths (6-18" and 18-30") are available for 2015 only (Research 
   Analytical Laboratory, UMN).

2. **Soil moisture in control plots** — EC-5 sensors were not installed in 
   control treatment plots (trt 21, 0 kg N) by design. VWC data unavailable 
   for 16 plots, preventing full assessment of moisture × N rate interactions.

3. **Lysimeter data interpretation** — Lysimeters measure NO3 concentration 
   in soil water at 100cm depth but do not provide drainage volume data 
   needed to calculate leaching rates. Plot-year means compress opposing 
   biological processes across the growing season. August 2014 is absent 
   due to harvest logistics. For these reasons formal statistical analysis 
   of lysimeter data is not performed in this document.

4. **Post-harvest soil NO3 data exclusion** — Two factors limit the 
   interpretability of the soil nitrate measurements: (a) lab methodology 
   differences between years — the 2014 Agvise lab reports values in rounded 
   increments (100% of 18-30" values at exactly 1.0 ppm; 84% of 6-18" at 
   1.0 ppm), representing a reporting threshold effect rather than biological 
   uniformity, while the 2015 RAL-UMN lab reports continuous decimal values 
   with 0.05 ppm resolution; and (b) post-harvest soil NO3 reflects the 
   cumulative outcome of multiple competing processes (plant uptake efficiency, 
   denitrification, leaching, microbial immobilization, fall mineralization) 
   with no way to separate their relative contributions. For these reasons, 
   formal statistical analysis of soil NO3 data is not performed in this 
   document.

5. **Nitrogen budget in intercrop plots** — The intercrop treatment receives 
   40 kg N ha⁻¹ of synthetic fertilizer plus variable additional nitrogen 
   from alfalfa biological fixation. Mineralization dynamics, alfalfa biomass, 
   and timing of N availability cannot be quantified from the available data.

6. **Multi-environment trial design** — The same 32 plots were measured in 
   both 2014 and 2015. Treatment effects are interpreted across four 
   environments (2 locations × 2 years) rather than as independent replicates.

---

## Exploratory Visualizations

### Grain Yield by Treatment, Location, and Year

Grain yield was compared across four treatments and two years. Waseca and 
Lamberton are shown as separate panels to reveal the site × year interaction. 
Error bars show ± 1 standard error (n = 4 plots per treatment per location).

```{r yield-bar-plot, fig.width=9, fig.height=5}
yield_summary <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    se_yield   = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

ggplot(yield_summary, aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Mean Kernza grain yield by treatment, location, and year",
    subtitle = "Error bars show ± 1 SE | n = 4 plots per treatment",
    x        = "Treatment",
    y        = "Mean grain yield (kg ha⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

Lamberton yields declined dramatically from 2014 to 2015 across all treatments 
while Waseca yields increased. This opposite directional response between sites 
demonstrates a strong treatment × environment interaction. Site-level 
differences in soil drainage and seasonal precipitation distribution likely 
drive the divergent yield responses.

---

### Grain Yield and Lodging by Treatment, Location, and Year

To evaluate whether lodging contributed to yield differences between years 
mean grain yield and lodging score are plotted together. If lodging caused 
yield decline, sites with higher lodging should show lower yield. Waseca shows 
the opposite pattern — yield increased from 2014 to 2015 despite increasing 
lodging scores — ruling out lodging as the primary cause of the Waseca yield 
improvement.

```{r yield-lodging-plot, fig.width=10, fig.height=8}
yield_lodging_plot <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield   = mean(grain_yield, na.rm = TRUE),
    se_yield     = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    mean_lodging = mean(lodging_score, na.rm = TRUE),
    se_lodging   = sd(lodging_score, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

p1 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Grain yield and lodging by treatment, location, and year",
       subtitle = "Waseca yield increased despite higher lodging — lodging not the cause",
       x        = NULL,
       y        = "Mean yield (kg ha⁻¹)",
       fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

p2 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_lodging, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_lodging - se_lodging,
                    ymax = mean_lodging + se_lodging),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(x    = "Treatment",
       y    = "Mean lodging score (0-10)",
       fill = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

p1 / p2
```

---

### Soil Water NO3 at 100cm Depth — Exploratory Only

Lysimeter NO3 concentrations (mg/L) from soil water samples at 100cm depth 
are shown below as exploratory context. Formal statistical analysis is NOT 
performed on this data due to the limitations noted above (single depth, 
concentration without flux, seasonal averaging, August 2014 gap, and high 
BDL frequency). Patterns suggest site-level differences between Waseca and 
Lamberton but should be interpreted cautiously.

```{r no3-plot, fig.width=9, fig.height=5}
ggplot(data, aes(x = trt, y = NO3_mean_mgL, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Soil water NO3 at 100cm depth by treatment, location, and year",
    subtitle = "Exploratory visualization only — not used for formal inference",
    x        = "Treatment",
    y        = "Mean NO3 concentration (mg L⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

---

### Soil Moisture Field Map

Seasonal mean volumetric water content (VWC) is displayed as a spatial 
field map by plot, treatment, depth, location, and year. Each cell 
represents one plot colored by its seasonal mean moisture. Grey cells 
labeled "NS" indicate no sensors were installed (by design) in the 
0 kg N control treatment.

```{r moisture-field-map, fig.width=12, fig.height=8}
moisture_field_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, vwc_100cm) %>%
  pivot_longer(cols = c(vwc_50cm, vwc_100cm),
               names_to  = "depth",
               values_to = "vwc") %>%
  mutate(
    depth = recode(depth,
                   "vwc_50cm"  = "50cm depth",
                   "vwc_100cm" = "100cm depth"),
    depth = factor(depth, levels = c("50cm depth", "100cm depth")),
    block = paste("Block", replicate)
  ) %>%
  group_by(location, year, depth, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup()

ggplot(moisture_field_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)), 
           fill = vwc)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(plot, "\n", 
                                ifelse(is.na(vwc), "NS", 
                                       round(vwc, 3)))),
            size = 2.3, lineheight = 0.9) +
  facet_grid(location ~ year + depth) +
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    values   = scales::rescale(c(0.13, 0.18, 0.22, 0.26, 0.30, 0.37)),
    limits   = c(0.13, 0.37),
    na.value = "grey80",
    name     = "VWC\n(m³/m³)",
    breaks   = c(0.15, 0.20, 0.25, 0.30, 0.35)) +
  labs(
    title    = "Soil moisture field map by plot location",
    subtitle = "Each cell = one plot | Grey cells (NS) = control plots with no sensors",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 9),
    axis.text        = element_text(size = 8),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    aspect.ratio     = 0.7)
```

---

### Spatial Relationship — Moisture, Yield, and Treatment

This three-layer visualization combines soil moisture (tile color), grain 
yield (circle size), and treatment (circle color) in the same spatial 
field layout. This allows simultaneous examination of moisture patterns, 
yield variation, and treatment effects at the plot level.

```{r moisture-yield-trt-map, fig.width=14, fig.height=9, dpi=150, out.width="100%"}
moisture_yield_trt_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, grain_yield) %>%
  mutate(block = paste("Block", replicate)) %>%
  group_by(location, year, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup() %>%
  filter(!is.na(grain_yield))

ggplot(moisture_yield_trt_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)))) +
  geom_tile(aes(fill = vwc_50cm), color = "white", linewidth = 1.5) +
  geom_point(aes(size = grain_yield, color = trt), 
             alpha = 0.85, stroke = 1.2) +
  geom_text(aes(label = plot), 
            size = 2, color = "black",
            fontface = "bold", vjust = 2.8) +
  facet_grid(location ~ year) +
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    limits   = c(0.13, 0.37),
    na.value = "grey85",
    name     = "VWC 50cm\n(m³/m³)") +
  scale_color_manual(values = c(
    "0 kg N"                 = "#2C3E50",
    "60 kg N"                = "#3498DB",
    "Kernza/Alfalfa 40 kg N" = "#27AE60",
    "120 kg N"               = "#C0392B"),
    name = "Treatment") +
  scale_size_continuous(
    range = c(3, 12),
    name  = "Grain yield\n(kg ha⁻¹)",
    breaks = c(50, 150, 300, 500)) +
  labs(
    title    = "Spatial relationship: soil moisture, yield, and treatment",
    subtitle = "Tile color = soil moisture | Circle size = yield | Circle color = treatment",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    legend.box       = "vertical") +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    color = guide_legend(order = 3, override.aes = list(size = 5)))
```

---

---

## ANOVA Research Questions

Packages for post-hoc comparisons are loaded here to avoid the MASS package 
masking `dplyr::select()` earlier in the document.

```{r load-anova-packages, message=FALSE, warning=FALSE}
library(emmeans)
library(multcomp)
library(multcompView)
```

---

### ANOVA Question 1: Does treatment affect grain yield and does this interact with environment?

**Hypothesis:** The four treatments (0 kg N, 60 kg N, 120 kg N, and 
Kernza/Alfalfa 40 kg N intercrop) will produce different grain yields, 
and this effect will vary by environment (location × year). Specifically:
- N fertilizer rate is expected to increase yield up to some saturation 
  point
- The intercrop treatment is hypothesized to outperform the 40 kg N 
  equivalent monoculture position through N fixation by alfalfa, with 
  the strongest benefit in year 3 (2014)
- The intercrop advantage may diminish by year 4 (2015) as competition 
  with the aging Kernza stand increases

```{r anova-q1, fig.width=10, fig.height=5}
model1 <- aov(grain_yield ~ trt * location * year, data = data)
summary(model1)

em1 <- emmeans(model1, ~ trt | location * year)
cld(em1, Letters = letters)

data %>%
  ggplot(aes(x = factor(year), y = grain_yield, fill = trt)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.85)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c(
    "0 kg N"                 = "#2C3E50",
    "60 kg N"                = "#3498DB",
    "Kernza/Alfalfa 40 kg N" = "#2ECC71",
    "120 kg N"               = "#C0392B")) +
  labs(title    = "Grain yield by treatment, location, and year",
       subtitle = "All four treatments shown | Intercrop = Kernza/Alfalfa 40 kg N",
       x        = "Year",
       y        = "Grain yield (kg ha⁻¹)",
       fill     = "Treatment") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"),
        legend.position  = "right")
```

**Interpretation:** The ANOVA revealed two highly significant effects: 
the main effect of treatment (p = 0.0037) and the location × year 
interaction (p = 0.0003). The significant treatment effect confirms 
that the four treatments produced meaningfully different grain yields 
across the dataset. The location × year interaction reflects the opposite 
directional yield response between sites — Waseca yields increased from 
2014 to 2015 while Lamberton yields decreased. The absence of a 
significant treatment × location or treatment × year interaction 
(p = 0.82 and p = 0.70 respectively) indicates that treatment effects 
were relatively consistent across environments, though the magnitude 
varied.

The boxplot reveals several important patterns: the 0 kg N control 
consistently produced the lowest yields in all environments; N fertilizer 
rate generally increased yield with some diminishing returns at 120 kg N; 
and the intercrop treatment (40 kg N synthetic + alfalfa N fixation) 
matched or exceeded the 120 kg N monoculture at Lamberton 2014, 
suggesting substantial alfalfa N contribution in the year 3 environment 
with highest yield potential. The intercrop advantage diminished in 
year 4 (2015) at both sites.

**Supplementary 2-level analysis — crop system effect:** For comparison, 
an additional ANOVA collapses the three N rate monoculture treatments 
(0, 60, 120 kg N) into a single "TLI monoculture" group to test whether 
the presence of a legume intercrop alone affects yield, independent of 
N rate variation.

```{r anova-q1-crops, fig.width=8, fig.height=5}
model1_crops <- aov(grain_yield ~ crops * location * year, data = data)
summary(model1_crops)

em1_crops <- emmeans(model1_crops, ~ crops | location * year)
cld(em1_crops, Letters = letters)

data %>%
  ggplot(aes(x = factor(year), y = grain_yield, fill = crops)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("TLI" = "#2E75B6",
                                "TLI/leg" = "#2ECC71")) +
  labs(title    = "Grain yield by crop system across environments",
       subtitle = "TLI = Kernza monoculture (pooled N rates) | TLI/leg = Kernza/alfalfa intercrop",
       x        = "Year",
       y        = "Grain yield (kg ha⁻¹)",
       fill     = "Crop system") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"))
```

Note that the crop system model shows a marginal effect (p = 0.073) 
compared to the highly significant full treatment model (p = 0.0037). 
The reduced statistical power reflects the fact that pooling the three 
N rates introduces substantial within-group variance, obscuring the 
clearer treatment-level effects visible in the 4-treatment analysis.

---

### ANOVA Question 2: Does plant height differ by treatment?

**Hypothesis:** Plant height will increase with nitrogen fertilizer rate 
as higher N availability promotes vegetative growth. The intercrop treatment 
may show different height dynamics due to competition with alfalfa.

```{r anova-q2, fig.width=8, fig.height=5}
model2 <- aov(plant_height ~ trt * year, data = data)
summary(model2)

em2 <- emmeans(model2, ~ trt)
cld(em2, Letters = letters)

ggplot(data, aes(x = trt, y = plant_height, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Plant height by treatment and year",
       subtitle = "Height measured from soil to top of seed head at physiological maturity",
       x        = "Treatment",
       y        = "Plant height (cm)",
       fill     = "Year") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

---

## Regression Research Questions

### Regression Question 1: Does grain yield increase with N fertilizer rate?

**Hypothesis:** Grain yield increases linearly with N fertilizer rate in 
Kernza monoculture plots. Based on Jungers et al. (2017) a quadratic 
relationship is also possible if yields decline at high N rates due to 
lodging. The linear model is tested first.

```{r regression-q1, fig.width=8, fig.height=5}
reg1 <- lm(grain_yield ~ n_fertilizer, data = data_mono)

summary(reg1)

ggplot(data_mono, aes(x = n_fertilizer, y = grain_yield)) +
  geom_point(aes(color = location, shape = year), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Kernza grain yield vs N fertilizer rate (monoculture only)",
    subtitle = "Linear regression | Points colored by location, shaped by year",
    x        = "N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year") +
  theme_bw()
```

---

### Regression Question 2: Where does the intercrop treatment fall on the yield ~ N rate curve?

**Hypothesis:** If the alfalfa intercrop provides a yield benefit through 
N fixation, the intercrop treatment (40 kg N ha⁻¹ synthetic + alfalfa-fixed N) 
should produce yields comparable to or exceeding monoculture plots receiving 
similar or higher synthetic N rates. If competition effects dominate, intercrop 
yields will fall below the monoculture regression line at 40 kg N ha⁻¹.

```{r regression-q2, fig.width=8, fig.height=5}
ggplot() +
  geom_smooth(data = data_mono,
              aes(x = n_fertilizer, y = grain_yield),
              method = "lm", se = TRUE,
              color = "#2E75B6", fill = "#2E75B6", alpha = 0.2) +
  geom_point(data = data_mono,
             aes(x = n_fertilizer, y = grain_yield),
             color = "#2E75B6", size = 2, alpha = 0.6) +
  geom_point(data = data %>% filter(crops == "TLI/leg"),
             aes(x = n_fertilizer, y = grain_yield),
             color = "#2ECC71", size = 4, shape = 17) +
  labs(
    title    = "Intercrop yield relative to monoculture N response curve",
    subtitle = "Blue = monoculture regression | Green triangles = intercrop",
    x        = "Synthetic N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)") +
  theme_bw()
```

---

### Regression Question 3: Does lodging score increase with plant height?

**Hypothesis:** Taller plants are more prone to lodging due to a higher 
center of gravity and increased wind resistance. However Jungers et al. 
(2017) found that plant height did not explain additional variation in 
lodging beyond the effect of N fertilizer rate, suggesting stem diameter 
or composition may be more important than height alone.

```{r regression-q3, fig.width=8, fig.height=5}
reg3 <- lm(lodging_score ~ plant_height, data = data)
summary(reg3)

ggplot(data, aes(x = plant_height, y = lodging_score)) +
  geom_point(aes(color = trt), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = c(
    "0 kg N"                 = "#D3D3D3",
    "60 kg N"                = "#74A9CF",
    "Kernza/Alfalfa 40 kg N" = "#2ECC71",
    "120 kg N"               = "#2C7BB6")) +
  labs(
    title    = "Lodging score vs plant height",
    subtitle = "Does plant height predict lodging susceptibility?",
    x        = "Plant height (cm)",
    y        = "Lodging score (0-10)",
    color    = "Treatment") +
  theme_bw()
```

---

### Regression Question 4: Does alfalfa legume biomass correlate with Kernza grain yield in intercrop plots?

**Hypothesis:** If the intercrop system primarily provides yield benefits 
through alfalfa N fixation, larger alfalfa biomass in an intercrop plot 
should positively correlate with Kernza grain yield in that same plot 
because more alfalfa biomass represents more N fixation potential. 
Alternatively, if competition for resources (water, light, nutrients) 
dominates the intercrop dynamic, alfalfa biomass may negatively correlate 
with Kernza yield. Testing this relationship addresses the mechanistic 
basis for the intercrop treatment's performance observed in Regression Q2.

**Statistical limitations:** This analysis uses only the 16 intercrop 
plots (n = 4 per environment × 4 environments). Statistical power is 
limited, and the location and year effects are expected to dominate 
the raw correlation. The regression below controls for location and year.

```{r regression-q4, fig.width=8, fig.height=5}
data_intercrop <- data %>% 
  filter(crops == "TLI/leg", !is.na(legume_biomass), !is.na(grain_yield))

reg4 <- lm(grain_yield ~ legume_biomass + location + year, 
           data = data_intercrop)
summary(reg4)

ggplot(data_intercrop, 
       aes(x = legume_biomass, y = grain_yield, 
           color = location, shape = factor(year))) +
  geom_point(size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, aes(group = location)) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Kernza grain yield vs alfalfa biomass in intercrop plots",
    subtitle = "Testing N fixation benefit vs resource competition",
    x        = "Alfalfa biomass (kg ha⁻¹)",
    y        = "Kernza grain yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year") +
  theme_bw()
```

**Interpretation:** At Lamberton, alfalfa biomass increased approximately 
4-fold from 2014 (mean 74 kg ha⁻¹) to 2015 (mean 323 kg ha⁻¹), while 
Kernza grain yield in the same plots declined approximately 3-fold (381 
to 131 kg ha⁻¹). At Waseca, alfalfa biomass remained low in both years 
due to poor establishment on the poorly drained soil. This inverse 
relationship at Lamberton suggests that in the year 3-to-4 transition, 
competition from established alfalfa may exceed the N fixation benefit — 
a finding consistent with recent literature showing intercrops can 
reduce Kernza yields in subsequent production years (Favre et al. 2024).

This result highlights the complexity of perennial intercrop systems 
and suggests future research is needed to identify management practices 
(planting density, timing, harvest strategies) that could optimize the 
balance between N fixation benefit and resource competition as stands age.

---

## Summary of Findings

```{r summary-table}
tibble(
  Question = c(
    "ANOVA Q1: Treatment × environment effect on grain yield",
    "ANOVA Q2: Treatment effect on plant height",
    "Regression Q1: Grain yield ~ N fertilizer rate (monoculture)",
    "Regression Q2: Intercrop position on yield curve",
    "Regression Q3: Lodging score ~ plant height",
    "Regression Q4: Kernza grain yield ~ alfalfa legume biomass (intercrop plots)"
  ),
  `Response Variable` = c(
    "grain_yield", "plant_height",
    "grain_yield", "grain_yield", "lodging_score", "grain_yield"
  )
) %>%
  kable(caption = "Summary of research questions and response variables")
```

---

## Future Directions

### Context: Documented June 2014 flooding at Waseca

The opposite yield trajectories between Waseca (2014→2015: +47% grain yield) 
and Lamberton (2014→2015: -49% to -66% grain yield across treatments) are 
the most striking pattern in this dataset. Weather records clarify a key 
driver at Waseca: June 2014 was the wettest month in Minnesota state 
history, with Waseca receiving 12.31 inches of rainfall — roughly four 
times the historical June average (Minnesota DNR 2014). Multiple flash 
flooding events affected Waseca County during the month, with flooding 
documented by National Weather Service alerts. The flooding occurred 
during the critical stem elongation and early reproductive development 
stages of Kernza on the poorly drained Webster clay loam soil, producing 
prolonged saturated and anaerobic conditions. The 2014 Waseca yields 
therefore represent suppressed yields under a documented extreme flooding 
event rather than typical year 3 Kernza performance.

The observation that 2015 Waseca grain yields nearly doubled while plant 
height, straw biomass, and lodging remained relatively stable suggests 
the crop expressed more of its reproductive potential once flooding stress 
was removed. This framing — combined with Lamberton's contrasting 2014-2015 
decline under normal conditions — raises testable questions about the 
mechanistic basis of year-over-year yield variation in perennial Kernza 
stands.

### Research priority 1: Does root biomass accumulation drive the year 
1-to-year 2 Kernza yield decline?

The well-documented year 1-to-year 2 grain yield decline in Kernza stands 
remains mechanistically unresolved and is identified as a key challenge 
for perennial grain crop development (Cattani & Asselin 2018; De Pauw 
et al. 2022). One plausible hypothesis emerges from established principles 
of perennial grass physiology.

Root maintenance respiration in perennial plants can consume over 50% of 
the carbon utilized within the root system (Eissenstat 1993). Because this 
maintenance cost scales with absolute root biomass, a larger year 2 Kernza 
root system would impose a substantially higher absolute carbon demand than 
a smaller year 1 system, even if proportional allocation rules remained 
unchanged. Similarly, nitrogen remobilization from vegetative tissues to 
developing grain (accounting for 65-90% of grain N in cereals; Hu et al. 
2016) must compete with N demands from an expanding root system.

Under this source-sink competition framework, the year 1-to-2 yield 
decline may result from increased absolute carbon and nitrogen consumption 
by root maintenance rather than from changes in allocation priorities. 
The plant's photosynthetic capacity remains similar, but a larger fraction 
of fixed photosynthate and cycled nitrogen is consumed by root maintenance 
before it becomes available for grain filling.

**Testable predictions:**

- Root biomass should increase substantially between year 1 and year 2 
  Kernza stands (already documented by Jungers and Gutknecht)
- Absolute carbon allocation to root maintenance respiration should 
  increase proportionally with root biomass
- Grain yield should correlate inversely with root maintenance carbon 
  demand across stand ages
- Nitrogen remobilization efficiency to grain should decrease as root 
  system N sink strength increases

**Methods:** Destructive root biomass sampling, root respiration 
measurements (chamber or isotope-tracer methods), whole-plant carbon 
allocation studies using 13C labeling, and plant tissue N analysis 
at key phenological stages. These methods are established in related 
perennial grass research (Carbone et al. 2007; Yao and Chen 2019) and 
transferable to Kernza.

**Status in literature:** No published study has directly tested this 
relationship in Kernza. Existing Kernza research documents root biomass 
increases with stand age but frames this as beneficial for soil carbon 
storage rather than as a potential yield-limiting cost. This represents 
a specific, tractable research gap.

### Research priority 2: What regulatory pathways control root biomass 
development in Kernza, and how does waterlogging stress alter them?

This priority is contingent on Priority 1 findings. If root biomass 
accumulation is confirmed as a significant contributor to yield decline, 
identifying regulatory pathways that control root biomass development 
could inform breeding targets for improved yield stability across years.

The 2014 Waseca flooding provides a natural experimental context. The 
flooding reduced grain yield while apparently limiting normal root 
expansion. Understanding the molecular basis of this response has 
implications for both site selection (Kernza is promoted for marginal 
sites that may include poorly drained areas) and breeding for 
waterlogging tolerance.

**Approach:** Controlled waterlogging experiments using established 
methods (container flooding, hydroponic hypoxia, rhizotrons; Barickman 
et al. 2019; Kong et al. 2022) could recreate the 2014 Waseca conditions 
in a controlled setting. Comparison of root growth, development, and 
architecture between stressed and unstressed plants would provide the 
phenotypic baseline. If phenotypic differences are substantial, targeted 
gene expression analysis could investigate candidate pathways including 
hypoxia response (ERF-VII transcription factors), root meristem 
regulation, and carbohydrate allocation signaling.

**Research sequence rationale:** Priority 1 must precede Priority 2 
because phenotypic characterization of the root-yield relationship 
provides the foundation for any molecular investigation. Investing in 
gene expression work before confirming the phenotypic relationship risks 
targeting the wrong mechanism.

### Additional data limitations and future research opportunities

Beyond the two priority research questions, several specific limitations 
of this dataset suggest additional future work:

- **Early-season N availability proxy** — May lysimeter NO3 concentrations 
  may indirectly indicate N moving past the plant root zone before plant 
  uptake. An inverse relationship was observed in Lamberton 2014 but 
  could not be confirmed across sites or years. Direct early-season soil 
  or plant tissue N measurements would rigorously test this relationship.

- **In-season moisture dynamics** — Seasonal mean VWC values cannot 
  resolve within-season stress periods or test moisture optimum 
  hypotheses. Higher-resolution moisture monitoring including control 
  plots would enable this testing.

- **Intercrop N budget quantification** — The intercrop treatment's 
  total N input (synthetic + alfalfa-fixed) could not be quantified 
  without additional measurements of alfalfa biomass, N fixation rates, 
  mineralization rates, and Kernza plant tissue N at key phenological 
  stages.

- **Leaching flux calculation** — True nitrate leaching rates require 
  concurrent drainage volume measurements at 100cm. Pairing lysimeter 
  concentrations with water flux measurements would enable direct N loss 
  estimation and strengthen site-level comparisons.

---

*Data wrangling and QA/QC fully documented in `data_wrangling.Rmd`.*  
*Reference: Jungers et al. 2017. Agron. J. 109:462–472.*
