---
title: "Kernza Data Wrangle Project — Analysis"
author: "Christopher Sherve"
date: 
output: html_document
---

## Setup

Packages are loaded and data is read from Google Sheets. The `emmeans`, 
`multcomp`, and `multcompView` packages are loaded later immediately before 
the ANOVA section to avoid the MASS package masking `dplyr::select()`.

```{r setup}
library(tidyverse)
library(googlesheets4)
library(lubridate)
library(patchwork)
library(knitr)
library(ggplot2)
# Resolve MASS package conflict with dplyr
select <- dplyr::select
filter <- dplyr::filter

# Load corrected dataset from Google Sheets
data <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1RZdL8GhR8bRt3f6G9pSMFs50jnjGvhdAajJhmgkZ0xs/edit?usp=sharing",
  sheet = "combined_clean_final_v2"
)
```

```{r factor-variables}
# Factor treatment, year, and location with readable labels
data <- data %>%
  mutate(
    trt = factor(trt,
                 levels = c(21, 22, 25, 23),
                 labels = c("0 kg N", "60 kg N",
                            "Kernza/Alfalfa 40 kg N", "120 kg N")),
    year     = factor(year),
    location = factor(location),
    # Create environment variable for T×E analysis
    environment = paste(location, year, sep = "_")
  )
```

```{r data-subsets, include=FALSE}
# Monoculture only — for yield ~ N rate regression
data_mono <- data %>% filter(crops == "TLI")

# 2015 only — for TOC and deeper soil depth analyses
data_2015 <- data %>% filter(year == "2015")

# Long format soil NO3 by depth — for ANOVA Q4
soil_long <- data %>%
  dplyr::select(year, location, plot, trt, crops, n_fertilizer,
                NO3_N_ppm_0_6, NO3_N_ppm_6_18, NO3_N_ppm_18_30) %>%
  pivot_longer(cols = c(NO3_N_ppm_0_6, NO3_N_ppm_6_18, NO3_N_ppm_18_30),
               names_to  = "depth",
               values_to = "NO3_soil_ppm") %>%
  mutate(depth = factor(recode(depth,
    "NO3_N_ppm_0_6"   = "0-6 inches",
    "NO3_N_ppm_6_18"  = "6-18 inches",
    "NO3_N_ppm_18_30" = "18-30 inches"),
    levels = c("0-6 inches", "6-18 inches", "18-30 inches"))) %>%
  filter(!is.na(NO3_soil_ppm))
```

---

## Data Overview

The primary dataset (`combined_clean_final_v2`) integrates yield, lysimeter 
nitrate, soil moisture, and soil chemistry data from Waseca and Lamberton, MN 
across 2014 and 2015 (Years 3 and 4 of the study). Full data cleaning and 
QA/QC steps are documented in `data_wrangling.Rmd`, `metadata_kernza.md`, 
and `lysimeter_QAQC_findings.md`.

```{r data-overview}
glimpse(data)
summary(data)
```

---

## Data Limitations

The following limitations were identified during data wrangling and QA/QC:

1. **Soil OM at depth** — Loss-on-ignition organic matter is only available 
   at 0-6 inch depth for 2014 . Deeper depths were available for 2015 only.
2. **Soil moisture** — EC-5 sensors were not installed in control treatment 
   plots (trt 21, 0 kg N). VWC data unavailable for 16 plots.
3. **Lysimeter seasonal averaging** — Plot-year means compress opposing 
   biological processes across the growing season. August 2014 gap coincides 
   with grain harvest period (mid-August; Jungers et al. 2017).
4. **Multi-environment trial design** — The same 32 plots were measured in 
   both 2014 and 2015. Treatment effects are interpreted across four 
   environments (2 locations × 2 years) rather than as independent replicates.

---

## Exploratory Visualizations

### Grain Yield by Treatment, Location, and Year

Grain yield was compared across four treatments and two years. Waseca and 
Lamberton are shown as separate panels to reveal the site × year interaction. 
Error bars show ± 1 standard error (n = 4 plots per treatment per location).

```{r yield-bar-plot, fig.width=9, fig.height=5}
yield_summary <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    se_yield   = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

ggplot(yield_summary, aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Mean Kernza grain yield by treatment, location, and year",
    subtitle = "Error bars show ± 1 SE | n = 4 plots per treatment",
    x        = "Treatment",
    y        = "Mean grain yield (kg ha⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

Lamberton yields declined dramatically from 2014 to 2015 across all treatments 
while Waseca yields increased. This opposite directional response between sites 
demonstrates a strong treatment × environment interaction. One possible explanation 
is with site-level differences in soil drainage and seasonal precipitation 
distribution driving divergent yield responses.

---

### Grain Yield and Lodging by Treatment, Location, and Year

To evaluate whether lodging contributed to yield differences between years, 
mean grain yield and lodging score are plotted together. If lodging caused 
yield decline, sites with higher lodging should show lower yield. Waseca shows 
the opposite pattern where yield increased from 2014 to 2015 despite increasing 
lodging scores, thus ruling out lodging as the primary cause of the Waseca yield 
improvement.

```{r yield-lodging-plot, fig.width=10, fig.height=8}
yield_lodging_plot <- data %>%
  group_by(location, trt, year) %>%
  summarise(
    mean_yield   = mean(grain_yield, na.rm = TRUE),
    se_yield     = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    mean_lodging = mean(lodging_score, na.rm = TRUE),
    se_lodging   = sd(lodging_score, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Panel 1 — Yield
p1 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_yield, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Grain yield and lodging by treatment, location, and year",
       subtitle = "Waseca yield increased despite higher lodging — lodging not the cause",
       x        = NULL,
       y        = "Mean yield (kg ha⁻¹)",
       fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

# Panel 2 — Lodging
p2 <- ggplot(yield_lodging_plot,
             aes(x = trt, y = mean_lodging, fill = year)) +
  geom_bar(stat = "identity", position = position_dodge(0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_lodging - se_lodging,
                    ymax = mean_lodging + se_lodging),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(x    = "Treatment",
       y    = "Mean lodging score (0-10)",
       fill = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))

p1 / p2
```

---

### Nitrate Leaching by Treatment, Location, and Year

Lysimeter NO3 concentrations (mg/L) are shown by treatment and location. 
Lamberton consistently shows higher leaching than Waseca across all treatments 
and years. Due to many missing data and low or below detected limits further analysis was skipped.   
BDL values are substituted at 0.05 mg/L (half detection limit).

```{r no3-plot, fig.width=9, fig.height=5}
ggplot(data, aes(x = trt, y = NO3_mean_mgL, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_wrap(~location) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(
    title    = "Lysimeter nitrate leaching by treatment, location, and year",
    subtitle = "BDL values substituted at 0.05 mg/L | Lamberton consistently higher than Waseca",
    x        = "Treatment",
    y        = "Mean NO3 concentration (mg L⁻¹)",
    fill     = "Year") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

---

### Soil Moisture by Depth and Location

Seasonal mean volumetric water content (VWC) is shown at 50cm and 100cm 
depth across years and locations. Note: trt 21 (0 kg N control) plots had 
no sensors installed — this is by design, not missing data.

```{r moisture-plot, fig.width=8, fig.height=5}
# Create field map layout for soil moisture
# Each location, year, and depth gets its own field map

moisture_field_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, vwc_100cm) %>%
  pivot_longer(cols = c(vwc_50cm, vwc_100cm),
               names_to  = "depth",
               values_to = "vwc") %>%
  mutate(
    depth = recode(depth,
                   "vwc_50cm"  = "50cm depth",
                   "vwc_100cm" = "100cm depth"),
    depth = factor(depth, levels = c("50cm depth", "100cm depth")),
    block = paste("Block", replicate),
    # Create a plot order within each block (1-8 position)
    # Based on the actual plot numbering sequence
  ) %>%
  group_by(location, year, depth, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup()

# Build the field map heat map — blocks stacked, plots across
ggplot(moisture_field_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)), 
           fill = vwc)) +
  geom_tile(color = "white", linewidth = 1.5) +
 geom_text(aes(label = paste0(plot, "\n", 
                              ifelse(is.na(vwc), "NS", 
                                     round(vwc, 3)))),
                      size = 2.3, lineheight = 0.9) +
  facet_grid(location ~ year + depth) +
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    values   = scales::rescale(c(0.13, 0.18, 0.22, 0.26, 0.30, 0.37)),
    limits   = c(0.13, 0.37),
    na.value = "grey80",
    name     = "VWC\n(m³/m³)",
    breaks   = c(0.15, 0.20, 0.25, 0.30, 0.35)) +
  labs(
    title    = "Soil moisture field map by plot location",
    subtitle = "Each cell = one plot | Grey cells = control plots (no sensors)",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 9),
    axis.text        = element_text(size = 8),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    aspect.ratio     = 0.7)
geom_text(aes(label = paste0(plot, "\n", 
                              ifelse(is.na(vwc), "NS", 
                                     round(vwc, 3)))),
          size = 2.3, lineheight = 0.9)
```

```{r}
# Prepare data combining all three variables
moisture_yield_trt_map <- data %>%
  dplyr::select(year, location, plot, replicate, trt, 
                vwc_50cm, grain_yield) %>%
  mutate(block = paste("Block", replicate)) %>%
  group_by(location, year, block) %>%
  arrange(plot) %>%
  mutate(plot_position = row_number()) %>%
  ungroup() %>%
  filter(!is.na(grain_yield))

ggplot(moisture_yield_trt_map,
       aes(x = factor(plot_position), 
           y = fct_rev(factor(block)))) +
  # Layer 1: Background tile showing moisture
  geom_tile(aes(fill = vwc_50cm), color = "white", linewidth = 1.5) +
  # Layer 2 & 3: Circle size = yield, circle color = treatment
  geom_point(aes(size = grain_yield, color = trt), 
             alpha = 0.85,
             stroke = 1.2) +
  # Plot number label below circle
  geom_text(aes(label = plot), 
            size = 2, 
            color = "black",
            fontface = "bold",
            vjust = 2.8) +
  facet_grid(location ~ year) +
  # Moisture fill scale (tile background)
  scale_fill_gradientn(
    colors   = c("#D35400", "#F39C12", "#F9E79F", 
                 "#A9DFBF", "#52BE80", "#1E8449"),
    limits   = c(0.13, 0.37),
    na.value = "grey85",
    name     = "VWC 50cm\n(m³/m³)") +
  # Treatment color scale (circle color)
  scale_color_manual(values = c(
    "0 kg N"                 = "#2C3E50",  # dark navy
    "60 kg N"                = "#3498DB",  # blue
    "Kernza/Alfalfa 40 kg N" = "#27AE60",  # green
    "120 kg N"               = "#C0392B"), # red
    name = "Treatment") +
  # Yield size scale (circle size)
  scale_size_continuous(
    range = c(3, 12),
    name  = "Grain yield\n(kg ha⁻¹)",
    breaks = c(50, 150, 300, 500)) +
  labs(
    title    = "Spatial relationship: soil moisture, yield, and treatment",
    subtitle = "Tile color = soil moisture | Circle size = yield | Circle color = treatment",
    x        = "Plot position within block",
    y        = "Block") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid       = element_blank(),
    panel.border     = element_rect(color = "black", linewidth = 0.5),
    legend.position  = "right",
    legend.box       = "vertical") +
  guides(
    fill  = guide_colorbar(order = 1),
    size  = guide_legend(order = 2),
    color = guide_legend(order = 3, override.aes = list(size = 5)))
```
### Waseca 2014 as a Documented Flooding Event

June 2014 was the wettest month in Minnesota state history (Minnesota DNR 
2014), with Waseca receiving 12.31 inches of rainfall.  This is approximately four times 
more than the historical June average. Multiple flash flooding events affected Waseca 
County during the month. Based on the timing of events the flooding occurred during the critical 
stem elongation and early reproductive development stages of Kernza. 
This critical physiological development stage combined with the Webster clay loam soil which poorly drains are possible reasons leading to the reduced yield in 2014 versus 2015 in Waseca.   
Under these abnormal conditions prolonged saturated and anaerobic 
soil conditions across the plot area negatively impact the soil 
microbial communities, soil invertebrates, and crop root function.

This context is essential for interpreting the 2014 Waseca yields. Rather 
than representing typical year 3 Kernza performance, the 2014 Waseca data 
likely represent suppressed yields under a documented extreme flooding 
event. The 2015 grain yields nearly doubling while plant 
height and straw biomass remained stable.  This response is consistent with the crop 
expressing its reproductive potential once the flooding stress was 
removed and soil conditions returned to normal.
---

### Post-Harvest Soil Nitrate by Depth and Location

Soil NO3 concentrations (ppm) measured in fall of each year across three depth 
increments are shown by location and treatment. Deeper depth data (6-18" and 
18-30") is available for both years since both labs measured NO3 at all depths. 
Organic matter at depth is only available for 2015.

```{r soil-no3-depth-plot, fig.width=9, fig.height=5}
ggplot(soil_long, aes(x = trt, y = NO3_soil_ppm, fill = location)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  facet_grid(depth ~ year) +
  scale_fill_manual(values = c("Waseca" = "#2E75B6",
                                "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Post-harvest soil nitrate by depth, location, and year",
    subtitle = "Soils sampled October 21, 2014 and fall 2015",
    x        = "Treatment",
    y        = "Soil NO3-N (ppm)",
    fill     = "Location") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"))
```

---

## ANOVA Research Questions

Packages for post-hoc comparisons are loaded here to avoid the MASS package 
masking `dplyr::select()` earlier in the document.

```{r load-anova-packages}
library(emmeans)
library(multcomp)
library(multcompView)
```

---

### ANOVA Question 1: Does crop system affect grain yield, and does this interact with environment?

**Hypothesis:** The intercrop system (TLI/leg) will produce different grain 
yields than Kernza monoculture (TLI), and this effect will vary by environment 
(location × year). Specifically, the intercrop is hypothesized to outperform 
monoculture in year 3 (2014) when alfalfa N fixation benefit is strongest, 
but this advantage may diminish by year 4 (2015) as competition increases 
with stand age.

```{r anova-q1, fig.width=8, fig.height=5}
# Full model with T×E interaction
model1 <- aov(grain_yield ~ crops * location * year, data = data)
summary(model1)

# Post-hoc comparisons by crop system
em1 <- emmeans(model1, ~ crops | location * year)
cld(em1, Letters = letters)

# Visualize crop system means by environment
data %>%
  group_by(crops, location, year) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    se_yield   = sd(grain_yield, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  ) %>%
  ggplot(aes(x = year, y = mean_yield, color = crops, group = crops)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = mean_yield - se_yield,
                    ymax = mean_yield + se_yield), width = 0.1) +
  facet_wrap(~location) +
  scale_color_manual(values = c("TLI" = "#2E75B6",
                                 "TLI/leg" = "#2ECC71")) +
  labs(title    = "Grain yield by crop system across environments",
       subtitle = "TLI = Kernza monoculture | TLI/leg = Kernza/alfalfa intercrop",
       x        = "Year",
       y        = "Mean grain yield (kg ha⁻¹)",
       color    = "Crop system") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"))
```

---

### ANOVA Question 3: Does plant height differ by treatment?

**Hypothesis:** Plant height will increase with nitrogen fertilizer rate, 
as higher N availability promotes vegetative growth. The intercrop treatment 
may show different height dynamics due to competition with alfalfa.

```{r anova-q3, fig.width=8, fig.height=5}
model3 <- aov(plant_height ~ trt * year, data = data)
summary(model3)

em3 <- emmeans(model3, ~ trt)
cld(em3, Letters = letters)

# Visualize
ggplot(data, aes(x = trt, y = plant_height, fill = year)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("2014" = "#2E75B6", "2015" = "#70AD47")) +
  labs(title    = "Plant height by treatment and year",
       subtitle = "Height measured from soil to top of seed head at physiological maturity",
       x        = "Treatment",
       y        = "Plant height (cm)",
       fill     = "Year") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

---

### ANOVA Question 4 (Bonus): Does post-harvest soil nitrate differ by location, treatment, and depth?

**Hypothesis:** Post-harvest soil nitrate concentrations will reflect the net 
balance of two competing processes: (1) physical nitrogen retention by clay 
mineral surfaces, which would favor higher soil NO3 at Waseca (Webster clay 
loam, high CEC); and (2) biological nitrogen immobilization by the active 
microbial community and living Kernza roots post-harvest, which could reduce 
measurable mineral NO3 at both sites. If biological immobilization dominates, 
residual soil NO3 may be similarly low at both sites despite fundamentally 
different leaching losses during the growing season.

```{r anova-q4, fig.width=9, fig.height=5}
model4 <- aov(NO3_soil_ppm ~ location * trt * depth, data = soil_long)
summary(model4)

em4 <- emmeans(model4, ~ location | depth)
cld(em4, Letters = letters)

# Visualize depth profile by location
soil_long %>%
  group_by(location, depth) %>%
  summarise(
    mean_NO3 = mean(NO3_soil_ppm, na.rm = TRUE),
    se_NO3   = sd(NO3_soil_ppm, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  ) %>%
  ggplot(aes(x = mean_NO3, y = depth, color = location)) +
  geom_point(size = 4) +
  geom_errorbar(aes(xmin = mean_NO3 - se_NO3,
                    xmax = mean_NO3 + se_NO3), width = 0.2) +
  scale_y_discrete(limits = rev) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(title    = "Post-harvest soil nitrate profile by location",
       subtitle = "Averaged across treatments and years | Error bars show ± 1 SE",
       x        = "Soil NO3-N (ppm)",
       y        = "Soil depth",
       color    = "Location") +
  theme_bw()
```

---

## Regression Research Questions

### Regression Question 1: Does grain yield increase with N fertilizer rate?

**Hypothesis:** Grain yield increases linearly with N fertilizer rate in 
Kernza monoculture plots. Based on Jungers et al. (2017), a quadratic 
relationship is also possible if yields decline at high N rates due to lodging. 
The linear model is tested first; if the relationship is not linear a quadratic 
model will be considered.

```{r regression-q1, fig.width=8, fig.height=5}
# Linear model — monoculture only
reg1 <- lm(grain_yield ~ n_fertilizer, data = data_mono)
summary(reg1)

# Plot with both sites and years visible
ggplot(data_mono, aes(x = n_fertilizer, y = grain_yield)) +
  geom_point(aes(color = location, shape = year), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  scale_color_manual(values = c("Waseca" = "#2E75B6",
                                 "Lamberton" = "#E74C3C")) +
  labs(
    title    = "Kernza grain yield vs N fertilizer rate (monoculture only)",
    subtitle = "Linear regression | Points colored by location, shaped by year",
    x        = "N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)",
    color    = "Location",
    shape    = "Year") +
  theme_bw()
```

---

### Regression Question 2: Where does the intercrop treatment fall on the yield ~ N rate curve?

**Hypothesis:** If the alfalfa intercrop provides a yield benefit through N 
fixation, the intercrop treatment (40 kg N ha⁻¹ synthetic + alfalfa-fixed N) 
should produce yields comparable to or exceeding monoculture plots receiving 
similar or higher synthetic N rates. If competition effects dominate, intercrop 
yields will fall below the monoculture regression line at 40 kg N ha⁻¹.

```{r regression-q2, fig.width=8, fig.height=5}
# Intercrop vs monoculture regression
ggplot() +
  # Monoculture regression line
  geom_smooth(data = data_mono,
              aes(x = n_fertilizer, y = grain_yield),
              method = "lm", se = TRUE,
              color = "#2E75B6", fill = "#2E75B6", alpha = 0.2) +
  # Monoculture points
  geom_point(data = data_mono,
             aes(x = n_fertilizer, y = grain_yield),
             color = "#2E75B6", size = 2, alpha = 0.6) +
  # Intercrop points highlighted
  geom_point(data = data %>% filter(crops == "TLI/leg"),
             aes(x = n_fertilizer, y = grain_yield),
             color = "#2ECC71", size = 4, shape = 17) +
  labs(
    title    = "Intercrop yield relative to monoculture N response curve",
    subtitle = "Blue = monoculture regression | Green triangles = intercrop (40 kg N + alfalfa N fixation)",
    x        = "Synthetic N fertilizer rate (kg N ha⁻¹)",
    y        = "Grain yield (kg ha⁻¹)") +
  theme_bw()
```

---

### Regression Question 4: Does lodging score increase with plant height?

**Hypothesis:** Taller plants are more prone to lodging due to a higher center 
of gravity and increased wind resistance. However, Jungers et al. (2017) found 
that plant height did not explain additional variation in lodging beyond the 
effect of N fertilizer rate, suggesting stem diameter or composition may be 
more important than height.

```{r regression-q4, fig.width=8, fig.height=5}
reg4 <- lm(lodging_score ~ plant_height, data = data)
summary(reg4)

ggplot(data, aes(x = plant_height, y = lodging_score)) +
  geom_point(aes(color = trt), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = c(
    "0 kg N"                 = "#D3D3D3",
    "60 kg N"                = "#74A9CF",
    "Kernza/Alfalfa 40 kg N" = "#2ECC71",
    "120 kg N"               = "#2C7BB6")) +
  labs(
    title    = "Lodging score vs plant height",
    subtitle = "Does plant height predict lodging susceptibility?",
    x        = "Plant height (cm)",
    y        = "Lodging score (0-10)",
    color    = "Treatment") +
  theme_bw()
```

---

## Summary of Findings

```{r summary-table}
# Quick summary of ANOVA p-values
tibble(
  Question = c(
    "ANOVA Q1: Crop system × environment effect on grain yield",
    "ANOVA Q2: Treatment × location effect on NO3 leaching",
    "ANOVA Q3: Treatment effect on plant height",
    "ANOVA Q4: Location × treatment × depth effect on soil NO3",
    "Regression Q1: Grain yield ~ N fertilizer rate (monoculture)",
    "Regression Q2: Intercrop position on yield curve",
    "Regression Q3: Soil surface NO3 ~ lysimeter NO3",
    "Regression Q4: Lodging score ~ plant height"
  ),
  `Key Variable` = c(
    "grain_yield", "NO3_mean_mgL", "plant_height", "NO3_soil_ppm",
    "grain_yield", "grain_yield", "NO3_mean_mgL", "lodging_score"
  )
) %>%
  kable(caption = "Summary of research questions and response variables")
```

---

*Data wrangling and QA/QC fully documented in `data_wrangling.Rmd`.*  
*Reference: Jungers et al. 2017. Agron. J. 109:462–472.*
